Discussion Draft

### 2020-04-03 doc/Discussion.rmd
cleaning up typical landscape section

The spindle based model proposed that sexual dimorphism in the recombination landscape is driven by difference strength in the spindle assembly checkpoint (Lane Kauppi). The relaxed selection imposed by an ineffective SAC results in a uniform distribution of positions (while maintaining positive interference).

Just as the difference in SAC (checkpoint) strength can lead to disinct evolutionary trajectories of the genome wide recombination rate for oocytes and spermatocytes, the other side of the same coin is that there is increased variance between cells for crossover number. (weaker / leaky checkpoint )

The telomeric placement of crossovers in the male recombiation landscape, minimizes the amount of sister cohesion connecting homologs. This conserved feature of the male recombination landscape, might act to synchornize the seperation homologs after alignment at metaphase, (and in the degargation of the sister cohesion connecting the homologs. Another conserved feature of gametogenesis is that the timing of spermatocyte metapphase is much shorter of time compared to oocytes.

Whereas, in males, the directional selection could cause decreased variance in the recombination landscape via telomeric positioning of crossovers. Under the telomeric bias  recombination landscape, entry into anaphase may be more synchronizes (and faster) given that the reductional separation of homologs is dependent on a minimal amount of sister cohesion connecting the homologs (Figure X).

>>>>

Under **spindle based model** we hypothesize that the sexual dimorphism in the recombination landsacpes hinges on sex differences on the requirements for chromosome cohesion in late meiosis I. <1> the irreversable process of the metaphase to anaphase transistion is initiated by the protyltic decay of the sister cohesion connecting homologs --along the tetrad arms (cite).

<2> The number and placement of crossovers alter the distribution of sister cohesion and the 'chromosome stucture' once kinetochores attached to MT-spindle and tension applied.  For single acrocentric chromosomes the sister cohesion distal to the crossover connects the homologs (when aligned on metaphase plate). This is known as cruciform tetrad is some systems.

This area is reduced when crossover position is shifted towards the telomere (ends of the chromsome).

The male crossover telomere bias could be adaptive in order to synchornize (or minimize) the timing of homologs sepearting at metaphase I.

<3> Conserved feature of gametogenesis -- is the length of timing.
faster spermatogenesis -- may select for synchronize the separation homologs or faster seperate after alignment at metaphase. While in oogenesis, the longer timing and arrests may select for overall more sister cohesion (a different homolog strucutres)

<In oocytes the slower cell cycle and prolonged arrest may require chromosome structures with greater stability on the MI spindle.>
